---
title: "COVID analysis"
author: "Jon Yearsley"
date: '`r format(Sys.Date(), " %d %B, %Y")`'
output: github_document
knit: (
  function(inputFile, encoding) { 

    pSubTitle <- 'README'

    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      params      = list(sub_title = pSubTitle),      
      output_file = pSubTitle) })
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#setwd("~/git_repos/corona/")

library(readxl)
library(ggplot2)
library(zoo)
library(tidyr)
#library(plotly)
```

A quick look at the European data on Covid-19.

Data downloaded from two sources

+ European Centre for Disease Prevention and Control (ECDPC). https://www.ecdc.europa.eu/en    

+ Johns Hopkins University Center for Systems Science and Engineering (CSSE), https://github.com/CSSEGISandData/COVID-19



## Select countries to visualise

```{r}
countryList = c('Ireland',
            'United Kingdom',
            'Sweden',
            'France',
            'Switzerland')
```



## Download data

Data from ECDPC

```{r}
download.file(url='https://www.ecdc.europa.eu/sites/default/files/documents/COVID-19-geographic-disbtribution-worldwide.xlsx',
              destfile='COVID-19-geographic-disbtribution-worldwide.xlsx')

d_ecdpc= read_excel("COVID-19-geographic-disbtribution-worldwide.xlsx")
d_ecdpc = subset(d_ecdpc, countriesAndTerritories%in%countryList)
d_ecdpc$country = as.factor(d_ecdpc$countriesAndTerritories)
d_ecdpc$julian = as.numeric(julian(as.Date(d_ecdpc$dateRep)))
d_ecdpc$ID = c(1:nrow(d_ecdpc))

# Order deaths by time
ind = order(d_ecdpc$julian, d_ecdpc$country) 
d_ecdpc= d_ecdpc[ind,]

d_ecdpc$cumdeaths = ave(d_ecdpc$deaths, 
                        d_ecdpc$country, 
                        FUN=function(x){cumsum(x)})

d_ecdpc$dataSource='ECDPC'
```


Data from CSSE

```{r}
url_csse = 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/'
cases_filename = 'time_series_covid19_confirmed_global.csv'
deaths_filename = 'time_series_covid19_deaths_global.csv'
d_cases = read.csv(url(paste0(url_csse, cases_filename)))
d_deaths = read.csv(url(paste0(url_csse, deaths_filename)))

tmp = droplevels(subset(d_cases, Country.Region%in%countryList))
d1 = pivot_longer(data=tmp, 
                 cols = c(5:ncol(tmp)), 
                 names_to='dateStr', 
                 values_to='cumcases')

tmp = droplevels(subset(d_deaths, Country.Region%in%countryList))
d2 = pivot_longer(data=tmp, 
                 cols = c(5:ncol(tmp)), 
                 names_to='dateStr', 
                 values_to='cumdeaths')

d_csse = aggregate(cbind(cumcases,cumdeaths) ~ dateStr + Country.Region, 
                    data=merge(d1,d2), 
                    FUN=sum, na.rm=TRUE)

d_csse$dateRep = as.POSIXct(strptime(d_csse$dateStr, format='X%m.%d.%y'))
d_csse$julian = as.numeric(julian(d_csse$dateRep))
d_csse$cumcases = as.double(d_csse$cumcases)
d_csse$cumdeaths = as.double(d_csse$cumdeaths)

vars = names(d_csse)
vars[2] = 'country'
names(d_csse) = vars

# Order deaths by time
ind = order(d_csse$julian, d_csse$country) 
d_csse= d_csse[ind,]

# Calculate number of new cases and number of new deaths
for (c in countryList) {
  tmp = subset(d_csse, country==c)
  if (nrow(tmp)>0) {
    tmp$cases = c(tmp$cumcases[1], diff(tmp$cumcases))
    tmp$deaths = c(tmp$cumdeaths[1], diff(tmp$cumdeaths))
    
    if (c==countryList[1]) {
      d_csse_final = tmp
    } else {
      d_csse_final = rbind(d_csse_final, tmp)
    }
  }
}

d_csse_final$dataSource='CSSE'
```

-------------

Find day when number of new cases exceeded X (where X is typically set at 50-300)
```{r}
casesThreshold = 50
casesStr = paste0('Days post ',casesThreshold,' cases') # String for graph labels

# Function to calculate days into epidemic
add_daysIn = function(d, threshold) {
  dplus = subset(d, cases>threshold)
  startDate = aggregate(julian~country, 
                        data=dplus, 
                        FUN=function(x) {min(x)})
  
  ind = match(dplus$country,
              startDate$country)
  dplus$daysIn = dplus$julian - startDate$julian[ind]
  return(dplus)
}

varKeep = c('dataSource','dateRep','country','julian',
            'cases','deaths','cumdeaths')


d2 = as.data.frame(rbind(add_daysIn(d_ecdpc[,varKeep], casesThreshold),
           add_daysIn(d_csse_final[,varKeep], casesThreshold)))

d2$dataSource = as.factor(d2$dataSource)
d2$dateRep = as.Date(d2$dateRep)
```



Calculate rolling average over a window (roughly 5- 7 days). One reason for doing this is because there's uncertainty in the dates due to the time it takes for deaths and cases to be registered, etc.
```{r}
dayWindow = 5
rollingStr = paste0('(',dayWindow,' day moving average)')  # String for graph labels
for (i in countryList) {
  for (s in c('CSSE','ECDPC')) {
    tmp=subset(d2, country==i & dataSource==s)
    tmp$deaths_rolling = rollmean(tmp$deaths, k=dayWindow, align='right', fill=NA)
    tmp$cumdeaths_rolling = rollmean(tmp$cumdeaths, k=dayWindow, align='right', fill=NA)
    tmp$cases_rolling = rollmean(tmp$cases, k=dayWindow, align='right', fill=NA)

    # # Calculate average 1 week before
    # tmp$cases_rolling = rollapply(tmp$cases, k=dayWindow, align='right', fill=NA)

        
    if (i==countryList[1] & s=='CSSE') {
      d3 = tmp
    } else {
      d3 = rbind(d3, tmp)
    }
  }
}
```


# Visualisations

### Tables of the latest new cases and deaths per country

```{r}
library(pander)
ind = aggregate(julian~dataSource+country, 
                data=d3, 
                FUN=max)

pander(d3[d3$julian%in%ind$julian,
          c('dataSource',
                   'country', 
                   'dateRep',
                   'daysIn',
                   'cases',
                   'deaths',
                   'cumdeaths')],
       col.names=c('Database',
                   'Country',
                   'Date',
                   paste('Days post\n',casesThreshold,'cases'),
                   'New \nCases',
                   'New \nDeaths',
                   'Total deaths'),
       caption='Table 1: The latest numbers from the European Centre for Diesease Prevention and Control (ECDPC). https://www.ecdc.europa.eu/en and Johns Hopkins University Center for Systems Science and Engineering (CSSE), https://github.com/CSSEGISandData/COVID-19',
       keep.line.breaks = TRUE,
       row.names=FALSE)
```


### Temporal trends

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,dataSource=='CSSE')

p1 = ggplot(data=d_plot,
       aes(x=daysIn, 
           y=cases_rolling, 
           colour=country)) +
  geom_point(na.rm=T) + 
  geom_smooth(method = 'loess', 
              formula=y~x,
              span=0.25,
              se=FALSE,
              na.rm=T) +
  scale_y_log10() +
  scale_color_brewer(palette = 'Dark2') +
  theme_bw() + 
  labs(x=casesStr,
       y=paste('Number of cases',rollingStr),
       title='Trend in new cases')

#ggplotly(p1)

p1
```

```{r message=FALSE, echo=FALSE}
p2=ggplot(data=d_plot,
       aes(x=daysIn, 
           y=deaths_rolling, 
           colour=country)) +
  geom_point(na.rm=T) + 
  geom_smooth(method = 'loess',
              formula=y~x,
              span=0.25,
              se=FALSE,
              na.rm=T) +
  scale_y_log10() +
  scale_color_brewer(palette = 'Dark2') +
  theme_bw() + 
  labs(x=casesStr,
       y=paste('Number of deaths',rollingStr),
       title='Trend in deaths')

#ggplotly(p2)
p2
```

```{r message=FALSE, echo=FALSE}
p3 = ggplot(data=d_plot,
       aes(y=deaths_rolling/cases_rolling, 
           x=daysIn, 
           colour=country)) +
  geom_point(na.rm=T) + 
  geom_smooth(method = 'loess', 
              span=0.75,
              se=FALSE,
              na.rm=T) +
  scale_color_brewer(palette = 'Dark2') +
  theme_bw() + 
  labs(y=paste('Deaths / Cases',rollingStr),
       x=casesStr,
       title='Crude death rate trend')

#ggplotly(p3)
p3
```

### Epidemic indicator

Calculate the deaths over `r dayWindow` days minus the same statistic in the preceeding window.

```{r}
d_plot$death_change = NA
d_plot$death_change[-c(1:dayWindow)] =   d_plot$deaths[-c(1:dayWindow)] 
```
... to be completed

```{r message=FALSE, echo=FALSE}
p4 = ggplot(data=d_plot,
       aes(x=cumdeaths_rolling, 
           y=deaths_rolling, 
           colour=country)) +
  geom_point(na.rm=T) + 
  geom_smooth(method = 'loess', 
              span=0.25,
              se=FALSE,
              na.rm=T) +
  scale_y_log10() +
  scale_x_log10() +
  scale_color_brewer(palette = 'Dark2') +
  theme_bw() + 
  labs(x=paste0('Cumulative number of deaths ',rollingStr),
       y=paste0('Number of deaths ',rollingStr),
       title='Cumulative verus current deaths')

#ggplotly(p4)
p4
```
