---
title: "COVID stats for Moran et al (2020) paper"
author: "Jon Yearsley"
date: '`r format(Sys.Date(), " %d %B, %Y")`'
output: github_document
knit: (
  function(inputFile, encoding) { 

    pSubTitle <- 'README'

    rmarkdown::render( 
      input       = inputFile, 
      encoding    = encoding, 
      params      = list(sub_title = pSubTitle),      
      output_file = pSubTitle) })
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#setwd("~/git_repos/corona/")

library(readxl)
library(ggplot2)
library(zoo)
library(tidyr)
library(plotly)
```

A quick look at the  data on Covid-19.

Data downloaded from 

+ Johns Hopkins University Center for Systems Science and Engineering (CSSE), https://github.com/CSSEGISandData/COVID-19



## Select countries to visualise

```{r}
countryList = c('Ireland',
            'United Kingdom',
            'Italy',
            'Spain',
            'France',
            'Germany',
            'Switzerland')
```



## Download data


Data from CSSE

```{r echo=FALSE}
url_csse = 'https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/'
cases_filename = 'time_series_covid19_confirmed_global.csv'
deaths_filename = 'time_series_covid19_deaths_global.csv'
d_cases = read.csv(url(paste0(url_csse, cases_filename)))
d_deaths = read.csv(url(paste0(url_csse, deaths_filename)))

tmp = droplevels(subset(d_cases, Country.Region%in%countryList))
d1 = pivot_longer(data=tmp, 
                 cols = c(5:ncol(tmp)), 
                 names_to='dateStr', 
                 values_to='cumcases')

tmp = droplevels(subset(d_deaths, Country.Region%in%countryList))
d2 = pivot_longer(data=tmp, 
                 cols = c(5:ncol(tmp)), 
                 names_to='dateStr', 
                 values_to='cumdeaths')

d_csse = aggregate(cbind(cumcases,cumdeaths) ~ dateStr + Country.Region, 
                    data=merge(d1,d2), 
                    FUN=sum, na.rm=TRUE)

d_csse$dateRep = as.POSIXct(strptime(d_csse$dateStr, format='X%m.%d.%y'))
d_csse$julian = as.numeric(julian(d_csse$dateRep, origin=as.Date("2020-01-22")))
d_csse$cumcases = as.double(d_csse$cumcases)
d_csse$cumdeaths = as.double(d_csse$cumdeaths)

vars = names(d_csse)
vars[2] = 'country'
names(d_csse) = vars

# Order deaths by time
ind = order(d_csse$julian, d_csse$country) 
d_csse= d_csse[ind,]

# Calculate number of new cases and number of new deaths
for (c in countryList) {
  tmp = subset(d_csse, country==c)
  if (nrow(tmp)>0) {
    tmp$cases = c(tmp$cumcases[1], diff(tmp$cumcases))
    tmp$deaths = c(tmp$cumdeaths[1], diff(tmp$cumdeaths))
    
    if (c==countryList[1]) {
      d_csse_final = tmp
    } else {
      d_csse_final = rbind(d_csse_final, tmp)
    }
  }
}

d_csse_final$dataSource='CSSE'
```

-------------

Find day when number of new cases exceeded X (where X is typically set at 50-300)
```{r echo=FALSE}
casesThreshold = 50
casesStr = paste0('Days post ',casesThreshold,' cases') # String for graph labels

# Function to calculate days into epidemic
add_daysIn = function(d, threshold) {
  dplus = subset(d, cases>threshold)
  startDate = aggregate(julian~country, 
                        data=dplus, 
                        FUN=function(x) {min(x)})
  
  ind = match(dplus$country,
              startDate$country)
  dplus$daysIn = dplus$julian - startDate$julian[ind]
  return(dplus)
}

varKeep = c('dataSource','dateRep','country','julian',
            'cases','deaths','cumdeaths')


d2 = as.data.frame(add_daysIn(d_csse_final[,varKeep], casesThreshold))

d2$dataSource = as.factor(d2$dataSource)
d2$dateRep = as.Date(d2$dateRep)
```



Calculate rolling average over a window (roughly 5- 7 days). One reason for doing this is because there's uncertainty in the dates due to the time it takes for deaths and cases to be registered, etc.
```{r echo=FALSE}
dayWindow = 5
rollingStr = paste0('(',dayWindow,' day moving average)')  # String for graph labels
for (i in countryList) {
  for (s in c('CSSE')) {
    tmp=subset(d2, country==i & dataSource==s)
    tmp$deaths_rolling = rollmean(tmp$deaths, k=dayWindow, align='right', fill=NA)
    tmp$cumdeaths_rolling = rollmean(tmp$cumdeaths, k=dayWindow, align='right', fill=NA)
    tmp$cases_rolling = rollmean(tmp$cases, k=dayWindow, align='right', fill=NA)

        
    if (i==countryList[1] & s=='CSSE') {
      d3 = tmp
    } else {
      d3 = rbind(d3, tmp)
    }
  }
}
```


# Visualisations

### Tables of stats at peak daily deaths (5 day mean)

```{r echo=FALSE}
library(pander)
ind = aggregate(deaths_rolling~country, 
                data=d3, 
                FUN=function(x) {max(x,na.rm=TRUE)})

peak_deaths = as.data.frame(ind)
peak_deaths$date = as.Date(NA)
peak_deaths$cases = NA
peak_deaths$cumdeaths = NA
peak_deaths$weeks = NA


for (i in 1:nrow(peak_deaths)) {
  logInd = which(d3$country==peak_deaths$country[i] & d3$deaths_rolling==peak_deaths$deaths_rolling[i])
  
  peak_deaths$date[i] = d3$dateRep[logInd] 
  peak_deaths$cases[i] = d3$cases[logInd] 
  peak_deaths$cumdeaths[i] = d3$cumdeaths[logInd] 
  peak_deaths$weeks[i] = d3$julian[logInd] /7
}

pander(peak_deaths,
       col.names=c('Country',
                   'Max \nDeaths',
                   'Date',
                   'New \nCases',
                   'Cum deaths',
                   'Weeks since 22nd Jan'),
       caption='Table 1: The numbers when deaths peaked. Data from  Johns Hopkins University Center for Systems Science and Engineering (CSSE), https://github.com/CSSEGISandData/COVID-19',
       keep.line.breaks = TRUE,
       row.names=FALSE)
```


### Temporal trends in daily cases

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Ireland')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='darkgreen') + 
 #xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (Ireland)')

#ggplotly(p1)
p1
```


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='United Kingdom')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='darkblue') + 
 # scale_y_log10() +
 #xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (UK)')

#ggplotly(p1)
p1
```
```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Italy')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='magenta') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (Italy)')

#ggplotly(p1)
p1
```


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Spain')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='darkorange') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (Spain)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='France')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='gold') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (France)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Germany')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='darkred')+
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (Germany)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Switzerland')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cases_rolling)) +
  geom_point(na.rm=T, colour='black') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y=paste('Number of cases',rollingStr),
       title='Trend in daily case rate (Switzerland)')

#ggplotly(p1)
p1
```

### Temporal trends in cumulative deaths


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Ireland')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='darkgreen') + 
 #xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (Ireland)')

#ggplotly(p1)
p1
```


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='United Kingdom')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='darkblue') + 
 # scale_y_log10() +
 #xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (UK)')

#ggplotly(p1)
p1
```


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Italy')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='magenta') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (Italy)')

#ggplotly(p1)
p1
```


```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Spain')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='darkorange') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (Spain)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='France')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='gold') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (France)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Germany')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='darkred')+
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (Germany)')

#ggplotly(p1)
p1
```

```{r , echo=FALSE}
# Select either ECDPC or CSSE data to plot
d_plot = subset(d3,country=='Switzerland')

p1 = ggplot(data=d_plot,
       aes(x=julian/7, 
           y=cumdeaths)) +
  geom_point(na.rm=T, colour='black') + 
 # scale_y_log10() +
# xlim(c(0,25)) +
  theme_bw() + 
  labs(x='Weeks since 22nd Jan',
       y='Number of cumulative deaths',
       title='Trend in cumulative deaths (Switzerland)')

#ggplotly(p1)
p1
```
